\chapter{Multi-Agent State Machines (DAGs \& Checkpointing)}

Production systems use \textbf{Directed Acyclic Graphs (DAGs)} to manage Multi-Agent state, ensuring that data flows in specific directions and defines rigid boundaries.

\section{The DAG Architecture (LangGraph)}

The following diagram shows the modular routing of Coder, Tester, and Reviewer states through a DAG.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    scale=0.8, transform shape,
    node distance=1.1cm and 1.5cm,
    state/.style={draw, ellipse, fill=blue!10, minimum height=0.7cm},
    agent/.style={draw, rectangle, rounded corners, fill=yellow!20, minimum height=0.7cm},
    arrow/.style={thick,->,>=stealth}
]

\node (start) [state] {Start State};
\node (router) [agent, below=of start] {Router Node (LLM)};
\node (coder) [agent, fill=red!20, below left=of router, xshift=-0.5cm] {Coder Agent};
\node (tester) [agent, fill=green!20, below right=of router, xshift=0.5cm] {Tester Agent};
\node (reviewer) [agent, fill=purple!20, below=of router, yshift=-1.2cm] {Reviewer Agent};
\node (deploy) [agent, fill=orange!20, below=of reviewer] {Deploy Agent};
\node (end) [state, below=of deploy] {End State};

\draw[arrow] (start) -- (router);
\draw[arrow] (router) -| node[above left, font=\footnotesize] {Write Code} (coder);
\draw[arrow] (router) -| node[above right, font=\footnotesize] {Write Tests} (tester);
\draw[arrow] (coder) |- (reviewer);
\draw[arrow] (tester) |- (reviewer);

\draw[arrow] (reviewer) -- node[right, font=\footnotesize] {Pass} (deploy);
\draw[arrow] (reviewer) -| node[above, font=\footnotesize] {Fail} (coder);
\draw[arrow] (deploy) -- (end);

\end{tikzpicture}
\caption{Multi-Agent Directed Acyclic Graph}
\end{figure}

\section{DAG Parallel Execution: Topological Sort}

If multiple agents do not share code dependencies, they should run in parallel to maximize throughput. To determine the execution sequence, the orchestrator computes a **Topological Sort** of the graph.

Here is a Python implementation of Kahn's Algorithm for resolving agent execution queues.


\begin{center}
\noindent\rule{0.6\textwidth}{0.4pt} \\
\vspace{0.3em}
\textit{[Code implementation is available in the companion coding handbook: \texttt{coding-handbook/ch10\_multi\_agent/}]} \\
\noindent\rule{0.6\textwidth}{0.4pt}
\end{center}


\section{Memory Checkpointing (Postgres)}

To pause and resume an agent workflow (for Human-in-the-loop approval), the global state must be serialized to a database (like PostgreSQL) after every node execution.


\begin{center}
\noindent\rule{0.6\textwidth}{0.4pt} \\
\vspace{0.3em}
\textit{[Code implementation is available in the companion coding handbook: \texttt{coding-handbook/ch10\_multi\_agent/}]} \\
\noindent\rule{0.6\textwidth}{0.4pt}
\end{center}


By pausing the graph at the \texttt{Deploy Agent} node, a human engineer can review the code, click ``Approve'', and call \texttt{resume\_workflow()}.

This concludes the Definitive Guide to Agentic AI. You have transitioned from prompt engineering to rigorous systems architecture.
